Live Functional Programming with Typed Holes

### 2018-02-06 calculus.tex
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\todo{exposition}

@@

\todo{definition of complete types and terms? do we care about completeness
  of $\hexp$s or $\dexp$s?}

@@

\inferrule[IAp]
{\isIndet{\dexp_1}\\
% \isFinal{\dexp_2}~\text{\todo{??}}}
 \isFinal{\dexp_2}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}

\inferrule[ICast]
{\isIndet{\dexp}}
{\isIndet{\dcast{\htau}{\dexp}}}

@@

\begin{theorem}[Expandability \todo{??}] ~
  \begin{enumerate}
    \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ for some $\dexp$ and $\Delta$.
    \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Correspondence \todo{??}] ~
  \begin{enumerate}
    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to
  consistency?}
\end{theorem}

\begin{theorem}[Complete Progress \todo{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ then either
$\isValue{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.

@@

\begin{theorem}[Complete Preservation \todo{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ and
$\stepsToD{\Delta}{\dexp}{\dexp'}$ then
$\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$
\end{theorem}

\begin{theorem}[Commutativity \todo{state this}]
\end{theorem}

@@

\todo{canonical forms}

\todo{indeterminate forms}

>>>>

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\cy{exposition}

@@

\cy{definition of complete types and terms? do we care about completeness
  of $\hexp$s or $\dexp$s?}

@@

\inferrule[IAp]
{\isIndet{\dexp_1}\\
% \isFinal{\dexp_2}~\text{\cy{??}}}
 \isFinal{\dexp_2}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}

%% \inferrule[ICast]
%% {\isIndet{\dexp}}
%% {\isIndet{\dcast{\htau}{\dexp}}}

@@

\begin{theorem}[Expandability \cy{??}] ~
  \begin{enumerate}
    \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ for some $\dexp$ and $\Delta$.
    \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Correspondence \cy{??}] ~
  \begin{enumerate}
    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{theorem}

@@

\begin{theorem}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \cy{or maybe up to
  consistency?}
\end{theorem}

\begin{theorem}[Complete Progress \cy{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ then either
$\isValue{\dexp}$ or $\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.

@@

\begin{theorem}[Complete Preservation \cy{??}]
If $\hcomplete{\hexp}$ and
$\expandSyn{\emptyset}{\hexp}{\htau}{\dexp}{\Delta}$ and
$\stepsToD{\Delta}{\dexp}{\dexp'}$ then
$\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$
\end{theorem}

\begin{theorem}[Commutativity \cy{state this}]
\end{theorem}

@@

\cy{canonical forms}

\cy{indeterminate forms}

### 2018-02-06 hazelnut-dynamics.tex
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\newcommand{\todo}[1]{{\color{red} #1}}

@@

\input{macros}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\todo{abstract}
\end{abstract}

@@

>>>>

\bibliographystyle{ACM-Reference-Format}

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{leftidx}
\usepackage{todonotes}

\usepackage{listings}%
\lstloadlanguages{ML}
\lstset{tabsize=2, 
basicstyle=\footnotesize\ttfamily, 
% keywordstyle=\sffamily,
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{purple},
mathescape=false,escapechar=\#,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML, showspaces=false,showstringspaces=false,xleftmargin=15pt, 
morekeywords={string, float, int},
classoffset=0,belowskip=\smallskipamount, aboveskip=\smallskipamount
}
\let\li\lstinline

@@

\input{commands}
\input{macros}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\cy{blurb from intro of grant below}

\noindent
Program editors are increasingly incorporating \emph{live programming
services}~\cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007,burckhardt2013s}, which interleave editing with
evaluation of the program.
%
However, live programming services typically cannot run incomplete
programs, or they raise an exception as soon as a hole is encountered
during evaluation.
%
%% When they % encounter an incomplete program, a common strategy is
%to simply % disable the live programming service, or revert to the
%last complete % edit state.
%
This deprives the programmer, and also other editor services, of
potentially valuable feedback about the dynamic behavior of the
visible program. Our second track of research will address this gap by defining a dynamic semantics for incomplete programs. We plan to focus the first year of research efforts largely on this track of research, as indicated in \autoref{fig:tracks}. 
% Evaluation of incomplete programs will proceed as far as possible past the hole, and in the process, will track the environment around each hole instance.
This semantics will support live programming services like that outlined in our example above, which provides the programmer with a result on every, or nearly every, edit and allows the programmer to choose a hole instance in the computed result of evaluation and inspect its environment (in the sidebar of \autoref{fig:intro-example}) while they determine how to best fill that hole. 
Tracking each hole instance's environment will also enhance the editor services specified by the remaining two tracks of research, \ie{}~the action suggestion panel (\trackThree{}) and the direct manipulation programming service (\trackFour{}).
% We will prove that in certain situations, hole filling commutes with evaluation.
% This formal observation will lead to a decrease in the cost of providing continuous feedback to the programmer on every edit by avoiding redundant recomputation.
Our preliminary work, described in a presentation at the \emph{LIVE 2017}
workshop~\cite{HazelLIVE}, has discovered encouraging connections between our proposed dynamics and previous research on contextual modal type theory (the Curry-Howard interpretation of contextual modal logic) \cite{Nanevski2008} and gradual type theory \cite{Siek06a}.

\end{abstract}

@@

\bibliography{references,all.short,hazel_NSF}

### 2018-02-06 intro.tex
wrote down major core metatheorems
pasted in relevant blurbs from LIVE2017 paper and grant to use as starting text

% !TEX root = hazelnut-dynamics.tex
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\todo{intro}

>>>>

% !TEX root = hazelnut-dynamics.tex
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\cy{intro from LIVE 2017 paper is below}

Broadly speaking, live programming environments are those that granularly interleave editing and evaluation \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90,McDirmid:2007:LUL:1297027.1297073,burckhardt2013s}. 
In the words of \citet{burckhardt2013s}, live programming environments 
``promise to narrow the temporal and perceptive gap 
between program development and code execution''. Examples of live programming environments include {lab notebook environments},
e.g. the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
programmer to interactively edit and evaluate program fragments organized into a
sequence of cells (an extension of the read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007:LUL:1297027.1297073}, Sketch-n-Sketch \cite{DBLP:conf/pldi/ChughHSA16} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples.

The problem that has motivated much of our recent work is that most  
programming environments, live programming environments included,  provide feedback via various editor services only once the program being edited is syntactically well-formed and, when relevant, well-typed. This leaves a ``temporal and perceptive gap'', because programmers sometimes leave a program malformed or ill-typed for extended periods of time, e.g. as they think about what to enter at the cursor, or as they work on a different part of the program.

In view of this general problem, we recently developed a \emph{structure editor calculus} called Hazelnut where every edit state consists of a well-formed and statically meaningful, i.e. well-typed, incomplete expression, which we take to mean an expression with holes \cite{popl-paper}. This calculus addressed fundamental questions relevant to editor services that operate statically, but there was no solution in that paper to the problems faced by editor services that also require knowledge of the dynamic meaning of an incomplete program, as would be relevant to a live programming environment. For example, consider a \emph{stepper} (a.k.a. a \emph{stepwise debugger}), like that available to Haskell programmers in the GHCi system \cite{GHC-stepper} and other systems \cite{DBLP:conf/haskell/MarlowIPG07,kar13566}, to OCaml programmers with recent work on \texttt{ocamli} by \citet{ocaml-stepper} and to Standard ML programmers \cite{DBLP:journals/jfp/TolmachA95}. A stepper requires that the expression being stepped be assigned dynamic meaning according to a small-step operational semantics \cite{DBLP:journals/jlp/Plotkin04a,pfpl}, but no such semantics was defined for incomplete expressions that arise when using Hazelnut, or any of these other systems. Defining such a semantics was left as future work in the Hazelnut paper, and in a subsequent ``vision paper'' \cite{snapl17-paper}. The purpose of this paper is to sketch out our progress toward a theoretically well-grounded solution.

\cy{intro from grant section is below}

Live programming environments granularly interleave editing and evaluation,
``promis[ing] to narrow the temporal and perceptive gap 
between program development and code execution''~\cite{burckhardt2013s}.
Examples of live programming environments include {lab notebook environments},
like the popular IPython/Jupyter~\cite{PER-GRA:2007}, which allow the
programmer to interactively edit and evaluate program fragments organized into a
sequence of cells (an extension of the ubiquitous read-eval-print loop (REPL)); spreadsheets; {live graphics programming environments} like SuperGlue \cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist} and the tools demonstrated by Bret Victor in his lectures \cite{victor2012inventing}; the TouchDevelop live UI framework \cite{burckhardt2013s}; live mobile application development systems like Flutter \cite{flutter}; and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, to name a few prominent examples. \cy{find somewhere to cite these tanimoto papers \cite{DBLP:conf/icse/Tanimoto13,DBLP:journals/vlc/Tanimoto90}}

Live programming environments typically cannot run incomplete programs, or they simply halt or raise an exception as soon as they encounter a hole during evaluation. This leaves a ``temporal and perceptive gap'' because the results of even those computations that do not depend critically on the hole are not available to the programmer. Our second track of research will address this gap by defining a dynamic semantics for incomplete programs that proceeds as far as possible around holes. 
Every program assigned static meaning by the semantics of Track 1 will also be assigned dynamic meaning by the semantics of Track 2.

For example, in \autoref{fig:intro-example} we saw that mapping the incomplete function \li{weighted_average} over the \li{grades} list produced an incomplete result. If holes simply raised exceptions, then the system would have aborted evaluation as soon as \li{weighted_average} was applied for the first time by \li{map}, and the programmer would not have been able to see that the result was a list of two elements, nor see the value of the sub-expression \li{10.0 *. g.hw1} for each element \li{g} of the \li{grades} list. Even programs with type inconsistencies can be evaluated under our proposed dynamics, because, as discussed in Track 1, non-empty holes operate as membranes around type inconsistencies. Taken together, the results from Tracks 1 and 2 may help to address a common complaint: that a static type discipline makes it difficult to do exploratory programming.

Our proposed dynamics goes beyond staged evaluation (\eg{}~\cite{Taha:1999}), partial evaluation (\eg{}~\cite{Jones:1993uq}) and symbolic evaluation (\eg{}~\cite{King:1976}) in that we track the dynamic environment (\ie{}~the substitutions performed) around each instance of an expression hole in the result. This allows the programmer to see, in the sidebar of \autoref{fig:intro-example}, the actual values that the variables in scope take on everywhere that the hole they are filling ends up during evaluation. This provides a hole-oriented specification of a standard feature of debuggers: inspection of the environment at designated points in the program. Finally, this environmental information is useful (though not required) for the edit action suggestion service of Track 3 and the direct manipulation programming service of Track 4.
